---
title: "exomel_fig3"
author: "Caroline Wheeler"
date: "9/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(randomForest)
library(ggplot2)
library(ROCR)
library(gdata)
```

load data
```{r}
# read in clinical data containing response
clin <- read.table("/fs/ess/PAS1695/projects/exorien/data/IO_NOVA_clin.txt", header = TRUE) %>%
  dplyr::rename("sample" = Sample_ID) %>%
  filter(Cancer_Type == "Melanoma") %>%
  arrange(sample)

response <- clin %>%
  select(sample, Response)

# read in relative abundance data
taxa <- read.csv("/fs/ess/PAS1695/projects/exorien/data/drake-output/2022-03-16/2022-03-16_RA-with-taxonomy.csv")

# selecting genus as it performs best
taxa.wide <- taxa %>%
  dplyr::select(sample, genus, exo.ra) %>%
  group_by(sample, genus) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'genus', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(taxa.wide$sample, clin$sample)

# didn't perform well
# read in prevalence data
# prev <- read_table("/fs/ess/PAS1695/projects/exorien/data/k2bout.txt") %>%
#     dplyr::filter(sample %in% clin$sample) %>%
#     arrange(sample)

# check sample order 
# identical(prev$sample, clin$sample)
  
# read in expression data - select 5 well perdforming expressions
expr <- read.csv("/fs/ess/PAS1695/projects/exorien/data/exomel_tmesig.csv") %>%
  select(sample, Rooney_Immune_Cytolytic, Cytotoxic, IFNg_18, Huang_NRS, TIP_Hot) %>%
  arrange(sample)

identical(expr$sample, clin$sample)

#  combined
# combo <- merge(taxa.wide, response)
```

######## Function to make the model and predictions ######
```{r}
getROC <- function(train, clin.train, test, clin.test){
  model.training <- randomForest(x = train[,-1, drop=FALSE], y = as.factor(clin.train$Response))
  test.preds <- predict(model.training, test[,-1, drop=FALSE])
  
  print(model.training)
  # Check variable importance
  varImpPlot(model.training)
  # Prediction confusion matrix
  table(observed = clin.test$Response, predicted = test.preds)
  
  prediction_for_roc_curve <- predict(model.training,test[,-1, drop=FALSE],type="prob")
  pred <- prediction(prediction_for_roc_curve[,2], clin.test$Response)
  perf <- performance(pred, "tpr", "fpr")
  
  auc_ROCR <- performance(pred, measure = "auc")
  print(auc_ROCR)
  auc <- auc_ROCR@y.values[[1]]
  print(auc)
  
  df <- data.frame(FalsePositive=c(perf@x.values[[1]]),
                    TruePositive=c(perf@y.values[[1]]))
  
  return(df)
}
```

### randomForest with RA on genus level ###

Split into testing and training data
```{r}
testn <- round(.2 * nrow(response))
set.seed(12345)
testsamps <- sample(response$sample, testn)

# exo with microbes
train.exo <- taxa.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.exo <- taxa.wide %>%
  dplyr::filter(sample %in% testsamps)

# all clin vars
clin.train <- response %>%
  dplyr::filter(!sample %in% testsamps)
clin.test <- response %>%
  dplyr::filter(sample %in% testsamps)
```

## model and plot ROC
```{r}
gen_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

gen_ra_plot <- ggplot(spec_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="darkgreen")

gen_ra_plot
```

## using 5 expressions 
```{r}
train.expr <- expr %>%
  dplyr::filter(!sample %in% testsamps)
test.expr <- expr %>%
  dplyr::filter(sample %in% testsamps)
```

```{r}
expr_roc <- getROC(train.expr, clin.train, test.expr, clin.test)

expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

expr_plot
```
### Genus + 5 expression, new model ###
```{r}
# merge
gen.expr <- merge(taxa.wide, expr)

# split in testing and training
train.gen.expr <- gen.expr %>%
  dplyr::filter(!sample %in% testsamps)
test.gen.expr <- gen.expr %>%
  dplyr::filter(sample %in% testsamps)

# model
gen.expr_roc <- getROC(train.gen.expr, clin.train, test.gen.expr, clin.test)
```
Plot combined ROC
```{r}
gen.expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

gen.expr_plot
```



Combined PCA + 5 expressions
THIS DOES NOT WORK - can only combine if same exact predictor variables? 
Error in randomForest::combine(rf_expr, rf_gen) :
Predictor variables are different in the randomForest objects.
PCA of genus level
# ```{r}
# gen.pca <- prcomp(gen.wide[,-1], center = TRUE, scale. = TRUE)
# gen.pca <- as.data.frame(gen.pca$x)
# gen.pca <- gen.pca[1:5]
# gen.pca$sample <- gen.wide$sample
# ```
# 
# ROC with first 5 PCA's
# ```{r}
# # exo with microbes
# train.gen.pca <- gen.pca %>%
#   dplyr::filter(!sample %in% testsamps)
# test.gen.pca <- gen.pca %>%
#   dplyr::filter(sample %in% testsamps)
# ```
# 
# model and plot ROC
# ```{r}
# pca_ra <- getROC(train.gen.pca, clin.train, test.gen.pca, clin.test)
# 
# pca_plot <- ggplot(pca_ra, aes(x=FalsePositive, y=TruePositive)) + 
#   theme_bw() +
#   geom_smooth(se = FALSE, color="darkgreen")
# 
# gen_ra_plot
# ```

### Combine 5 expressions and microbe PCA
# ```{r}
# rf_expr <- randomForest(x = train.expr[,-1, drop=FALSE], y = as.factor(clin.train$Response))
# rf_gen <- randomForest(x = train.gen.pca[,-1, drop=FALSE], y = as.factor(clin.train$Response))
# 
# rf_combined <- randomForest::combine(rf_expr, rf_gen)
# ```




##### Tested but not using #######

## model and plot ROC
```{r}
spec_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

spec_ra_plot <- ggplot(spec_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

spec_ra_plot
```


### RandomForest with RA on family level ###
```{r}
fam.wide <- taxa %>%
  dplyr::select(sample, family, exo.ra) %>%
  group_by(sample, family) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'family', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(fam.wide$sample, clin$sample)
```

Split fam RA into testing and training data
```{r}
# exo with microbes
train.fam <- fam.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.fam <- fam.wide %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
fam_ra <- getROC(train.fam, clin.train, test.fam, clin.test)

fam_ra_plot <- ggplot(fam_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

fam_ra_plot
```


### RandomForest with prevalence on species level ###
Split genus RA into testing and training data
```{r}
# exo with microbes
train.prev <- prev %>%
  dplyr::filter(!sample %in% testsamps)
test.prev <- prev %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
prev_roc <- getROC(train.prev, clin.train, test.prev, clin.test)

prev_plot <- ggplot(prev_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

prev_plot
```

### RandomForest with Roh Immune Score ###
```{r}
train.expr <- expr %>%
  dplyr::filter(!sample %in% testsamps)
test.expr <- expr %>%
  dplyr::filter(sample %in% testsamps)
```

```{r}
expr_roc <- getROC(train.expr, clin.train, test.expr, clin.test)

expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

expr_plot
```

### RandomForest with RA order ###
```{r}
ord.wide <- taxa %>%
  dplyr::select(sample, order, exo.ra) %>%
  group_by(sample, order) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'order', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(ord.wide$sample, clin$sample)
```

Split fam RA into testing and training data
```{r}
# exo with microbes
train.ord <- ord.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.ord <- ord.wide %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
ord_ra <- getROC(train.ord, clin.train, test.ord, clin.test)

ord_ra_plot <- ggplot(ord_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

ord_ra_plot
```

#### Roh Immune Score + Genus #####
```{r}
rf_expr <- randomForest(x = train.expr[,-1, drop=FALSE], y = as.factor(clin.train$Response))
rf_gen <- randomForest(x = train.gen[,-1, drop=FALSE], y = as.factor(clin.train$Response))

rf_combined <- randomForest::combine(rf_expr, rf_gen)
```

### Layer ROC curves ###
```{r}
temp <- combine(combo_ra, ord_ra, expr_roc, prev_roc, spec_ra, gen_ra, fam_ra)

combined <- ggplot(temp, aes(x=FalsePositive, y=TruePositive, color=source)) + 
  theme_bw() +
  geom_smooth(se = FALSE)

combined

ggsave(filename = "../figures/combined_ROC.png", plot = combined, width = 28, height = 10, dpi = 400)
```

Try other expressions
```{r}
expr <- read.csv("/fs/ess/PAS1695/projects/exorien/data/exomel_tmesig.csv") %>%
  select(sample, Ipi_neoadjuvant) %>%
  arrange(sample)

train.expr <- expr %>%
  dplyr::filter(!sample %in% testsamps)
test.expr <- expr %>%
  dplyr::filter(sample %in% testsamps)

expr_roc <- getROC(train.expr, clin.train, test.expr, clin.test)

expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_line()
  geom_smooth(se = FALSE, color="red3")

expr_plot
```
