---
title: "exomel_fig3"
author: "Caroline Wheeler"
date: "9/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(randomForest)
library(ggplot2)
#library(pROC)
library(ROCR)
library(gdata)
```

load data
```{r}
# read in clinical data containing response
clin <- read.table("/fs/ess/PAS1695/projects/exorien/data/IO_NOVA_clin.txt", header = TRUE) %>%
  dplyr::rename("sample" = Sample_ID) %>%
  filter(Cancer_Type == "Melanoma") %>%
  arrange(sample)

response <- clin %>%
  select(sample, Response)

# read in relative abundance data
taxa <- read.csv("/fs/ess/PAS1695/projects/exorien/data/drake-output/2022-03-16/2022-03-16_RA-with-taxonomy.csv")
  
taxa.wide <- taxa %>%
  dplyr::select(sample, microbe, exo.ra) %>%
  spread(key = 'microbe', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(taxa.wide$sample, clin$sample)

# read in prevalence data
prev <- read_table("/fs/ess/PAS1695/projects/exorien/data/k2bout.txt") %>%
    dplyr::filter(sample %in% clin$sample) %>%
    arrange(sample)

# check sample order 
identical(prev$sample, clin$sample)
  
# read in expression data - select 'Roh Immune Score'
expr <- read.csv("/fs/ess/PAS1695/projects/exorien/data/exomel_tmesig.csv") %>%
  select(sample, Roh_Immune_Score) %>%
  arrange(sample)

identical(expr$sample, clin$sample)

#  combined
# combo <- merge(taxa.wide, response)
```

######## Functions to make the model and predictions ######
```{r}
getROC <- function(train, clin.train, test, clin.test){
  model.training <- randomForest(x = train[,-1, drop=FALSE], y = as.factor(clin.train$Response))
  test.preds <- predict(model.training, test[,-1, drop=FALSE])
  
  print(model.training)
  # Check variable importance
  varImpPlot(model.training)
  # Prediction confusion matrix
  table(observed = clin.test$Response, predicted = test.preds)
  
  prediction_for_roc_curve <- predict(model.training,test[,-1, drop=FALSE],type="prob")
  pred <- prediction(prediction_for_roc_curve[,2], clin.test$Response)
  perf <- performance(pred, "tpr", "fpr")
  
  auc_ROCR <- performance(pred, measure = "auc")
  print(auc_ROCR)
  auc <- auc_ROCR@y.values[[1]]
  print(auc)
  
  df <- data.frame(FalsePositive=c(perf@x.values[[1]]),
                    TruePositive=c(perf@y.values[[1]]))
  
  return(df)
}
```

### randomForest with RA on species level ###

Split into testing and training data
```{r}
testn <- round(.2 * nrow(response))
set.seed(12345)
testsamps <- sample(response$sample, testn)

# exo with microbes
train.exo <- taxa.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.exo <- taxa.wide %>%
  dplyr::filter(sample %in% testsamps)

# all clin vars
clin.train <- response %>%
  dplyr::filter(!sample %in% testsamps)
clin.test <- response %>%
  dplyr::filter(sample %in% testsamps)
```

## model and plot ROC
```{r}
spec_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

spec_ra_plot <- ggplot(spec_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

spec_ra_plot
```


### RandomForest with RA on genus level ###

filter taxa to genus level
```{r}
gen.wide <- taxa %>%
  dplyr::select(sample, genus, exo.ra) %>%
  group_by(sample, genus) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'genus', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(gen.wide$sample, clin$sample)
```

Split genus RA into testing and training data
```{r}
# exo with microbes
train.gen <- gen.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.gen <- gen.wide %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
gen_ra <- getROC(train.gen, clin.train, test.gen, clin.test)

gen_ra_plot <- ggplot(gen_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="darkgreen")

gen_ra_plot
```

### RandomForest with RA on family level ###
```{r}
fam.wide <- taxa %>%
  dplyr::select(sample, family, exo.ra) %>%
  group_by(sample, family) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'family', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(fam.wide$sample, clin$sample)
```

Split fam RA into testing and training data
```{r}
# exo with microbes
train.fam <- fam.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.fam <- fam.wide %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
fam_ra <- getROC(train.fam, clin.train, test.fam, clin.test)

fam_ra_plot <- ggplot(fam_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

fam_ra_plot
```



### RandomForest with prevalence on species level ###
Split genus RA into testing and training data
```{r}
# exo with microbes
train.prev <- prev %>%
  dplyr::filter(!sample %in% testsamps)
test.prev <- prev %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
prev_roc <- getROC(train.prev, clin.train, test.prev, clin.test)

prev_plot <- ggplot(prev_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

prev_plot
```
### RandomForest with Roh Immune Score ###
```{r}
train.expr <- expr %>%
  dplyr::filter(!sample %in% testsamps)
test.expr <- expr %>%
  dplyr::filter(sample %in% testsamps)
```

```{r}
expr_roc <- getROC(train.expr, clin.train, test.expr, clin.test)

expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

expr_plot
```

### RandomForest with RA order ###
```{r}
ord.wide <- taxa %>%
  dplyr::select(sample, order, exo.ra) %>%
  group_by(sample, order) %>%
  summarize(exo.ra = sum(exo.ra)) %>%
  spread(key = 'order', value = 'exo.ra') %>%
  dplyr::filter(sample %in% clin$sample) %>%
  arrange(sample)

# check that RA and clin data line up
identical(ord.wide$sample, clin$sample)
```

Split fam RA into testing and training data
```{r}
# exo with microbes
train.ord <- ord.wide %>%
  dplyr::filter(!sample %in% testsamps)
test.ord <- ord.wide %>%
  dplyr::filter(sample %in% testsamps)
```

model and plot ROC
```{r}
ord_ra <- getROC(train.ord, clin.train, test.ord, clin.test)

ord_ra_plot <- ggplot(ord_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="blue")

ord_ra_plot
```

#### Roh Immune Score + Genus #####
```{r}
rf_expr <- randomForest(x = train.expr[,-1, drop=FALSE], y = as.factor(clin.train$Response))
rf_gen <- randomForest(x = train.gen[,-1, drop=FALSE], y = as.factor(clin.train$Response))

rf_combined <- randomForest::combine(rf_expr, rf_gen)
```

```{r}

```

### Layer ROC curves ###
```{r}
temp <- combine(combo_ra, ord_ra, expr_roc, prev_roc, spec_ra, gen_ra, fam_ra)

combined <- ggplot(temp, aes(x=FalsePositive, y=TruePositive, color=source)) + 
  theme_bw() +
  geom_smooth(se = FALSE)

combined

ggsave(filename = "../figures/combined_ROC.png", plot = combined, width = 28, height = 10, dpi = 400)
```




Try other expressions
```{r}
expr <- read.csv("/fs/ess/PAS1695/projects/exorien/data/exomel_tmesig.csv") %>%
  select(sample, MitoScore) %>%
  arrange(sample)

train.expr <- expr %>%
  dplyr::filter(!sample %in% testsamps)
test.expr <- expr %>%
  dplyr::filter(sample %in% testsamps)

expr_roc <- getROC(train.expr, clin.train, test.expr, clin.test)

expr_plot <- ggplot(expr_roc, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="red3")

expr_plot
```


```{r}
data(iris)
rf1 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)
rf2 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)
rf3 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)
rf.all <- randomForest::combine(rf1, rf2, rf3)
print(rf.all)
```



