---
title: "Figure4_redo"
author: "Caroline Wheeler"
date: "12/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Figure 4 #

This file generates the single sample GSEA results and the effect
size results as seen in figure 4.

```{r Required Packges, include=FALSE}
library(tidyverse)
library(ComplexHeatmap)
library(GSVA)
library(ggpubr)
library(Hmisc)
```

### Effect Size Plot ###

**Required Data**
- Microbe Data

```{r Load and format required data}
data <- readRDS("../data/deseqResults_redo.RDS")
taxa <- read.csv("/fs/ess/PAS1695/projects/exorien-melio/data/humanRNAfilt_counts_with_taxonomy.csv")

taxa <- taxa %>%
  select(domain, kingdom, phylum, class, order, family, genus, species) %>%
  pivot_longer(!c(phylum), names_to = "Taxonomic Level", values_to = "delabel") %>%
  distinct()

data <- data %>%
  mutate(delabel = rownames(data)) %>%
  left_join(taxa) %>%
  filter(threshold == TRUE) %>%
  mutate(delabel = substr(delabel, 4, nchar(delabel)),
         "Taxonomic Grouping" = ifelse(is.na(phylum), delabel, substr(phylum, 4, nchar(phylum))),
         "Taxonomic Grouping" = ifelse(`Taxonomic Grouping` %in% c("Actinobacteria", "Apicomplexa", "Ascomycota", "Basidiomycota", "Firmicutes", "Proteobacteria", "Uroviricota"), `Taxonomic Grouping`, "Other")) %>%
  column_to_rownames(var = "delabel")
```

```{r Generate Microbe Order and Effect Size}
microbe_order <- data %>%
  dplyr::arrange(desc(log2FoldChange)) %>%
  rownames(.) 

data$`Taxonomic Grouping` <- factor(data$`Taxonomic Grouping`, levels = c( "Actinobacteria", "Proteobacteria", "Uroviricota", "Other"))

Microbe_Effect_Size <- data %>%
  ggplot(aes(x=log2FoldChange, y=reorder(rownames(data), log2FoldChange), fill=`Taxonomic Grouping`)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  ylab(" ") +
  xlab("Effect Size") +
  scale_fill_manual(values=c("red3","cornflowerblue", "deeppink4", "darkslateblue")) +
  theme(legend.key.size = unit(1, 'cm'), legend.title = element_text(size=10), legend.text = element_text(size=8), axis.text =element_text(size = 10), axis.title=element_text(size=10))

Microbe_Effect_Size

# ggsave(Microbe_Effect_Size, filename = "../figures/fig1_effectSize_redo.png", width = 8, height = 6)
```

### Single single Gene Set Enrichment Analysis ###

**Required Data**  
- IO_NOVA Melanoma Expression Data  
- IO_NOVA Melanoma Clinical Data  
- Microbe Data  
- Hallmark Gene Set  
- TIMEx Cell Types  
- IOSig Gene Set
```{r Load in required data, include=FALSE}
seq_data <- readr::read_tsv("../data/IO_NOVA_mel_seq.txt")
clin_data <- readr::read_tsv("../data/IO_NOVA_mel_clin.txt")
microbe_data <- read.csv("../data/microbes_redo.csv") %>%
  dplyr::rename("sample" = X)
names(microbe_data) <- gsub(x = names(microbe_data), pattern = "\\.", replacement = " ")  
load("../data/Hallmark.rda")
load("../data/TIMEx.rda")
load("../data/IOSig_geneset.rda")
seq_data <- seq_data[!duplicated(seq_data$Genes), ]
```

```{r Generate Expression Set Object}
genExpSet <- function(seq, clin){
    # Prep Seq #
  seq <- as.data.frame(seq)
  genes <- seq$Genes
  seq <- seq[, -1]
  rownames(seq) <- genes
  
  # Prep Clin #
  clin <- as.data.frame(clin)
  sample_ids <- clin$Sample_ID
  rownames(clin) <- sample_ids
  
  seq <- seq[, match(sample_ids, colnames(seq))]
  
  seq <- seq %>%
    dplyr::mutate(Row_sum = rowSums(.)) %>%
    dplyr::filter(Row_sum != 0) %>%
    dplyr::select(-Row_sum)
    
  seq <- as.matrix(seq)
  
  clin <- new("AnnotatedDataFrame", data = clin)
  out <- Biobase::ExpressionSet(assayData = seq, phenoData = clin)
  out
}

ExpSet <- genExpSet(seq_data, clin_data)
```

```{r Run ssGSEA, include=FALSE}
# run single sample GSEA on the Hallmark, TIMEx, and IOSig signatures gene sets #
Hallmark_results <- GSVA::gsva(ExpSet, gset.idx.list = Hallmark, method = 'ssgsea')
Hallmark_results <- Biobase::exprs(Hallmark_results)
GeneSig_results <- GSVA::gsva(ExpSet, gset.idx.list = IOSig_geneset, method = 'ssgsea')
GeneSig_results <- Biobase::exprs(GeneSig_results)
TIMEx_results <- GSVA::gsva(ExpSet, gset.idx.list = TIMEx, method = 'ssgsea')
TIMEx_results <- Biobase::exprs(TIMEx_results)
```

```{r Combining Microbe and ssGSEA data, warning=FALSE}
combineDat <- function(gene_set, microbe){
  gene_set <- as.data.frame(gene_set)
  gene_set$Pathway <- rownames(gene_set)
  gene_set <- gene_set %>%
    tidyr::pivot_longer(cols = -Pathway, names_to = "sample", values_to = "NES")
  
  microbe_data <- microbe_data %>%
    tidyr::pivot_longer(cols = -sample, names_to = "Microbe", values_to = "Count")
  out <- dplyr::inner_join(gene_set, microbe_data, by = "sample")
  out <- out %>%
    tidyr::pivot_wider(names_from = Pathway, values_from = NES) %>%
    tidyr::pivot_wider(names_from = Microbe, values_from = Count)
  out
}

# Combine Microbe data with the ssGSEA results for clustering #
Hall_Micro <- combineDat(Hallmark_results, microbe = microbe_data)
GeneSig_Micro <- combineDat(GeneSig_results, microbe = microbe_data)
TIMEx_Micro <- combineDat(TIMEx_results, microbe = microbe_data)
```


```{r Generate Clustered Heatmaps}
clustHeat <- function(dat, gene_set, microbeDat, microbeOrder, title, x.text){
  
  Cor_Dat <- dat[, -1]
  Cor_Dat <- as.matrix(Cor_Dat)
  Cor_Dat <- Hmisc::rcorr(Cor_Dat, type = "spearman")
  
  gs_labels <- names(gene_set)

  Filtered_Dat <- reshape2::melt(Cor_Dat$r) %>%
    dplyr::rename(Gene_Set = Var1, Microbe = Var2) %>%
    dplyr::filter(Gene_Set %in% gs_labels,
                  Microbe %in% colnames(microbeDat)[-1])
  
  Filtered_Dat$Gene_Set <- stringr::str_remove_all(Filtered_Dat$Gene_Set, "HALLMARK_")
  
   GS <- tidyr::pivot_wider(Filtered_Dat, names_from = "Gene_Set", values_from = "value")
   microbe_names <- GS[, 1]
   GS <- as.matrix(GS[, -1]) # -1 to omit microbe name column from matrix
   rownames(GS) <- microbe_names$Microbe
  
  # Cluster based on euclidean distance
  clust <- hclust(dist(t(GS)))
  
  ggplot(Filtered_Dat, aes(x=Gene_Set, y=Microbe, fill=value)) +
   geom_tile(color = "black") +
    labs(title = title) +
      scale_fill_gradient2(low = "#075AFF",
                         mid = "#FFFFCC",
                         high = "#FF0000") +
    coord_fixed() +
    theme(axis.text.x = element_text(angle = 45, hjust=1, size = x.text),
          axis.text.y = element_text(size = 10),
          legend.position = 'bottom') +
    scale_x_discrete(limits = colnames(GS)[clust$order]) +
    scale_y_discrete(limits = rev(microbeOrder))
}

# Plot the 
Hall_clust <- clustHeat(Hall_Micro, Hallmark, microbe_data, microbe_order, "Hallmark Gene Sets", x.text = 6)
TIMEx_clust <- clustHeat(TIMEx_Micro, TIMEx, microbe_data, microbe_order, "TIMEx Cell Types", x.text = 8)
IOSig_clust <- clustHeat(GeneSig_Micro, IOSig_geneset, microbe_data, microbe_order, "IOSig Signatures", x.text = 8)
```

### Hallmark 
```{r}
temp <- Hall_clust$data
```

Set gene label angles 
```{r}
labs <- c(colnames(Hall_clust$plot_env$GS)[Hall_clust$plot_env$clust$order]) 

lab.df <- as.data.frame(cbind(x = 1:50, labs))

lab.df$ang <- seq(from=(120/nrow(lab.df))/1.5, to=(1.5* (120/nrow(lab.df)))-120, length.out=nrow(lab.df)) + 80
lab.df$hjust <- 0
lab.df$hjust[which(lab.df$ang < -90)] <- 1
lab.df$ang[which(lab.df$ang < -90)] <- (180+lab.df$ang)[which(lab.df$ang < -90)]
```

```{r}
heat <- temp %>%
  ggplot(aes(x=as.numeric(fct_relevel(Gene_Set, colnames(Hall_clust$plot_env$GS)[Hall_clust$plot_env$clust$order])), y=as.numeric(fct_relevel(Microbe, rev(microbe_order))), fill=value)) +
  geom_tile(show.legend = FALSE, color = "black") +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  geom_text(data = lab.df, inherit.aes = FALSE, aes(x = as.numeric(x), y = 17, label = labs, angle=ang, hjust=hjust), size=1.2) + 
  ylim(c(-40,35)) +
  xlim(c(0, 150)) +
  theme_void() +
  coord_polar() 

heat
ggsave("../figures/hall_clust_heatmap_redo.pdf", dev = "pdf")
```
### IOSig 
```{r}
temp <- IOSig_clust$data
```

set gene label angles
```{r}
labs <- c(colnames(IOSig_clust$plot_env$GS)[IOSig_clust$plot_env$clust$order]) 

labs.filler <- as.data.frame(cbind(x=1:55, labs=NA))
lab.df <- as.data.frame(cbind(x = 56:86, labs))
lab.df <- rbind(labs.filler, lab.df)

lab.df$ang <- seq(from=(208.8/nrow(lab.df))/1.5, to=(1.5* (208.8/nrow(lab.df)))-208.8, length.out=nrow(lab.df)) + 80
lab.df$hjust <- 0
lab.df$hjust[which(lab.df$ang < -90)] <- 1
lab.df$ang[which(lab.df$ang < -90)] <- (180+lab.df$ang)[which(lab.df$ang < -90)]
```


```{r}
temp %>%
  ggplot(aes(x=as.numeric(fct_relevel(Gene_Set, colnames(IOSig_clust$plot_env$GS)[IOSig_clust$plot_env$clust$order])) + 55, y=as.numeric(fct_relevel(Microbe, rev(microbe_order))), fill=value)) +
   geom_tile(show.legend = FALSE, color = "black") +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  geom_text(data = lab.df, inherit.aes = FALSE, aes(x = as.numeric(x), y = 17, label = labs, angle=ang, hjust=hjust), size=1.2) + 
  ylim(c(-40,35)) +
  xlim(c(0, 150)) +
  theme_void() +
  coord_polar()

ggsave("../figures/iosig_heatmap_redo.pdf", dev = "pdf")
```

### Timex
```{r}
temp <- TIMEx_clust$data
```

set gene label angles
```{r}
labs <- c(colnames(TIMEx_clust$plot_env$GS)[TIMEx_clust$plot_env$clust$order]) 

# FOR GRANT VERSION
#labs <- sub(".*_", "", labs)  
# END FOR GRANT VERSION

labs.filler <- as.data.frame(cbind(x=1:91, labs=NA))
lab.df <- as.data.frame(cbind(x = 92:128, labs))
lab.df <- rbind(labs.filler, lab.df)

lab.df$ang <- seq(from=(307.2/nrow(lab.df))/1.5, to=(1.5* (307.2/nrow(lab.df)))-307.2, length.out=nrow(lab.df)) + 80
lab.df$hjust <- 0
lab.df$hjust[which(lab.df$ang < -90)] <- 1
lab.df$ang[which(lab.df$ang < -90)] <- (180+lab.df$ang)[which(lab.df$ang < -90)]
```


```{r}
temp %>%
  ggplot(aes(x=as.numeric(fct_relevel(Gene_Set, colnames(TIMEx_clust$plot_env$GS)[TIMEx_clust$plot_env$clust$order])) + 91, y=as.numeric(fct_relevel(Microbe, rev(microbe_order))), fill=value)) +
   geom_tile(show.legend = FALSE, color = "black") +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  geom_text(data = lab.df, inherit.aes = FALSE, aes(x = as.numeric(x), y = 17, label = labs, angle=ang, hjust=hjust), size=1.2) + 
  ylim(c(-40,35)) +
  xlim(c(0, 150)) +
  theme_void() +
  coord_polar()

ggsave("../figures/timex_clust_heatmap_redo.pdf", dev = "pdf")
```

# ```{R Generate P-Value Matricies}
# getPvals <- function(dat, gene_set, microbeDat, microbeOrder, gene_set_order){
#   Cor_Dat <- dat[, -1]
#   Cor_Dat <- as.matrix(Cor_Dat)
#   Cor_Dat <- Hmisc::rcorr(Cor_Dat, type = "spearman")
#   
#   gs_labels <- names(gene_set)
# 
#   Filtered_Dat <- reshape2::melt(Cor_Dat$P) %>%
#     dplyr::rename(Gene_Set = Var1, Microbe = Var2) %>%
#     dplyr::filter(Gene_Set %in% gs_labels,
#                   Microbe %in% colnames(microbeDat)[-1])
#   
#   Filtered_Dat$Gene_Set <- stringr::str_remove_all(Filtered_Dat$Gene_Set, "HALLMARK_")
#   
#    GS <- tidyr::pivot_wider(Filtered_Dat, names_from = "Gene_Set", values_from = "value")
#    microbe_names <- GS[, 1]
#    GS <- as.matrix(GS[, -1]) # -1 to omit microbe name column from matrix
#    rownames(GS) <- microbe_names$Microbe
#    
#    pValueMatrix <- GS[microbeOrder, gene_set_order]
#    
#    return(pValueMatrix)
# }
# 
# Hall_PVal <- getPvals(Hall_Micro, Hallmark, microbe_data, microbe_order, gene_set_order = Hall_clust[["plot_env"]][["clust"]][["order"]])
# TIMEx_PVal <- getPvals(TIMEx_Micro, TIMEx, microbe_data, microbe_order, gene_set_order = TIMEx_clust[["plot_env"]][["clust"]][["order"]])
# IOSig_PVal <- getPvals(GeneSig_Micro, IOSig_geneset, microbe_data, microbe_order, gene_set_order = IOSig_clust[["plot_env"]][["clust"]][["order"]])
# ```

```{r Combine Plots}
clustered_plots <- ggarrange(Hall_clust,
                             TIMEx_clust,
                             IOSig_clust,
                             nrow = 3, ncol = 1, common.legend = T)

Combined_Plots <- ggarrange(clustered_plots, Microbe_Effect_Size, ncol = 2)

Combined_Plots
```
