---
title: "circle heatmap"
author: "Caroline Wheeler"
date: "3/24/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Figure 4 #

This file generates the single sample GSEA results and the effect
size results as seen in figure 4.

```{r Required Packges, include=FALSE}
library(tidyverse)
library(ComplexHeatmap)
library(GSVA)
#library(ggpubr)
```


### Effect Size Plot ###

**Required Data**
- Microbe Data

```{r Load and format required data}
data <- readRDS("../data/deseqResults.RDS")
data <- data %>%
  filter(level == "species" & !is.na(delabel))

new.rownames <- substr(rownames(data), 4, nchar(rownames(data)))
rownames(data) <- new.rownames
```

```{r Generate Microbe Order and Effect Size}
microbe_order <- data %>%
  dplyr::arrange(desc(log2FoldChange)) %>%
  rownames(.)

Microbe_Effect_Size <- data %>%
  ggplot(aes(x=log2FoldChange, y=reorder(rownames(data), log2FoldChange), fill=`Taxonomic Grouping`)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  ylab("Species") +
  xlab("Effect Size") +
  scale_fill_manual(values=c("red3", "cyan4", "darkgreen", "darkgoldenrod2", "salmon2", "cornflowerblue", "deeppink4")) +
  theme(legend.key.size = unit(1, 'cm'), legend.title = element_text(size=10), legend.text = element_text(size=8), axis.text =element_text(size = 10), axis.title=element_text(size=10))

Microbe_Effect_Size
```

### Single single Gene Set Enrichment Analysis ###

**Required Data**  
- IO_NOVA Melanoma Expression Data  
- IO_NOVA Melanoma Clinical Data  
- Microbe Data  
- Hallmark Gene Set  
- TIMEx Cell Types  
- IOSig Gene Set
```{r Load in required data, include=FALSE}
seq_data <- readr::read_tsv("../data/IO_NOVA_mel_seq.txt")
clin_data <- readr::read_tsv("../data/IO_NOVA_mel_clin.txt")
microbe_data <- readr::read_csv("../data/Microbes.csv")
load("../data/Hallmark.rda")
load("../data/TIMEx.rda")
load("../data/IOSig_geneset.rda")
seq_data <- seq_data[!duplicated(seq_data$Genes), ]
```

```{r Generate Expression Set Object}
genExpSet <- function(seq, clin){
    # Prep Seq #
  seq <- as.data.frame(seq)
  genes <- seq$Genes
  seq <- seq[, -1]
  rownames(seq) <- genes
  
  # Prep Clin #
  clin <- as.data.frame(clin)
  sample_ids <- clin$Sample_ID
  rownames(clin) <- sample_ids
  
  seq <- seq[, match(sample_ids, colnames(seq))]
  
  seq <- seq %>%
    dplyr::mutate(Row_sum = rowSums(.)) %>%
    dplyr::filter(Row_sum != 0) %>%
    dplyr::select(-Row_sum)
    
  seq <- as.matrix(seq)
  
  clin <- new("AnnotatedDataFrame", data = clin)
  out <- Biobase::ExpressionSet(assayData = seq, phenoData = clin)
  out
}

ExpSet <- genExpSet(seq_data, clin_data)
```

```{r Run ssGSEA, include=FALSE}
# run single sample GSEA on the Hallmark, TIMEx, and IOSig signatures gene sets #
Hallmark_results <- GSVA::gsva(ExpSet, gset.idx.list = Hallmark, method = 'ssgsea')
Hallmark_results <- Biobase::exprs(Hallmark_results)
GeneSig_results <- GSVA::gsva(ExpSet, gset.idx.list = IOSig_geneset, method = 'ssgsea')
GeneSig_results <- Biobase::exprs(GeneSig_results)
TIMEx_results <- GSVA::gsva(ExpSet, gset.idx.list = TIMEx, method = 'ssgsea')
TIMEx_results <- Biobase::exprs(TIMEx_results)
```

```{r Combining Microbe and ssGSEA data}
combineDat <- function(gene_set, microbe){
  gene_set <- as.data.frame(gene_set)
  gene_set$Pathway <- rownames(gene_set)
  gene_set <- gene_set %>%
    tidyr::pivot_longer(cols = -Pathway, names_to = "Sample_ID", values_to = "NES")
  
  microbe_data <- microbe_data %>%
    tidyr::pivot_longer(cols = -Sample_ID, names_to = "Microbe", values_to = "Prevalence")
  out <- dplyr::inner_join(gene_set, microbe_data, by = "Sample_ID")
  out <- out %>%
    tidyr::pivot_wider(names_from = Pathway, values_from = NES) %>%
    tidyr::pivot_wider(names_from = Microbe, values_from = Prevalence)
  out
}

# Combine Microbe data with the ssGSEA results for clustering #
Hall_Micro <- combineDat(Hallmark_results, microbe = microbe_data)
GeneSig_Micro <- combineDat(GeneSig_results, microbe = microbe_data)
TIMEx_Micro <- combineDat(TIMEx_results, microbe = microbe_data)
```

```{r}
  ggplot(aes(x=as.numeric(as.factor(Gene_Set)), y=as.numeric(Microbe), fill=value)) +
  geom_tile(show.legend = FALSE, color = "black") +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  geom_text(data = lab.df, inherit.aes = FALSE, aes(x = as.numeric(x), y = 70, label = labs, angle=ang, hjust=hjust), size=1.2) + 
  ylim(c(10,90)) +
  xlim(c(0, 150)) +
  theme_void() +
  coord_polar() 
```

Hall_clust plot
```{r}
Cor_Dat <- cor(Hall_Micro[, -1], method = "spearman")

gs_labels <- names(Hallmark)

Filtered_Dat <- reshape2::melt(Cor_Dat) %>%
  dplyr::rename(Gene_Set = Var1, Microbe = Var2) %>%
  dplyr::filter(Gene_Set %in% gs_labels,
                Microbe %in% colnames(microbe_data)[-1])

Filtered_Dat$Gene_Set <- stringr::str_remove_all(Filtered_Dat$Gene_Set, "HALLMARK_")

GS <- tidyr::pivot_wider(Filtered_Dat, names_from = "Gene_Set", values_from = "value")
GS <- as.matrix(GS[, -1]) # -1 to omit categories from matrix

# Cluster based on euclidean distance
clust <- hclust(dist(t(GS)))

ggplot(Filtered_Dat, aes(x=as.numeric(as.factor(Gene_Set)), y=as.numeric(Microbe), fill=value)) +
 geom_tile(color = "black") +
    scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  theme(axis.text.x = element_text(angle = 45, hjust=1, size = 6),
        axis.text.y = element_text(size = 10),
        legend.position = 'bottom') +
  scale_x_discrete(limits = as.numeric(as.factor(colnames(GS)[clust$order]))) +
  scale_y_discrete(limits = rev(microbe_order)) +
  theme_void() +
  coord_fixed()
```

TIMEx plot
```{r}

```

IOSig plot
```{r}

```

old version
```{r}
clustHeat <- function(dat, gene_set, microbeDat, microbeOrder, title, x.text){
  Cor_Dat <- cor(dat[, -1], method = "spearman")

gs_labels <- names(gene_set)

Filtered_Dat <- reshape2::melt(Cor_Dat) %>%
  dplyr::rename(Gene_Set = Var1, Microbe = Var2) %>%
  dplyr::filter(Gene_Set %in% gs_labels,
                Microbe %in% colnames(microbeDat)[-1])

Filtered_Dat$Gene_Set <- stringr::str_remove_all(Filtered_Dat$Gene_Set, "HALLMARK_")

 GS <- tidyr::pivot_wider(Filtered_Dat, names_from = "Gene_Set", values_from = "value")
 GS <- as.matrix(GS[, -1]) # -1 to omit categories from matrix

# Cluster based on euclidean distance
clust <- hclust(dist(t(GS)))

ggplot(Filtered_Dat, aes(x=Gene_Set, y=Microbe, fill=value)) +
 geom_tile(color = "black") +
  labs(title = title) +
    scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, size = x.text),
        axis.text.y = element_text(size = 10),
        legend.position = 'bottom') +
  scale_x_discrete(limits = colnames(GS)[clust$order]) +
  scale_y_discrete(limits = rev(microbeOrder))
}

# Plot the 
Hall_clust <- clustHeat(Hall_Micro, Hallmark, microbe_data, microbe_order, "Hallmark Gene Sets", x.text = 6)
TIMEx_clust <- clustHeat(TIMEx_Micro, TIMEx, microbe_data, microbe_order, "TIMEx Cell Types", x.text = 8)
IOSig_clust <- clustHeat(GeneSig_Micro, IOSig_geneset, microbe_data, microbe_order, "IOSig Signatures", x.text = 8)
```

