---
title: "microSig"
author: "Caroline Wheeler"
date: "2022-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(randomForest)
library(ggplot2)
library(ROCR)
library(gdata)
```

read in data
```{r}
# read in metmel sigs
sigs <- read.csv("../data/signatures-filtered-metastatic-melanoma_20221207.csv")

response <- read.table("/fs/ess/PAS1695/projects/exorien/data/IO_NOVA_clin.txt", header = TRUE) %>%
  dplyr::rename("sample" = Sample_ID) %>%
  filter(Cancer_Type == "Melanoma") %>%
  dplyr::select(sample, Response) 

# read in microbe data from exorien-melio 
taxa_all <- read.csv("/fs/ess/PAS1695/projects/exorien/data/drake-output/2022-03-16/2022-03-16_RA-with-taxonomy.csv") %>%
  filter(sample %in% response$sample)
  #mutate(Taxonomy = ifelse(grepl("d__Bacteria", Taxonomy), sub('d','k',Taxonomy), Taxonomy))

taxa <- taxa_all %>%
  select(sample, species, exo.ra) %>%
  rename(microbe = species)
```

get relAbun at all levels and combine into one table
```{r}
# @level is a string containing the taxonomic level we want to calculate relative abundances for
get_ra <- function(level){
  df <- taxa_all %>%
    group_by(sample, !!! rlang::syms(level)) %>%
    summarize(exo.ra = sum(exo.ra)) %>%
    rename(microbe = level)
  
  return(df)
}

dom <- get_ra("domain")
king <- get_ra("kingdom")
phyl <- get_ra("phylum")
class <- get_ra("class")
ord<- get_ra("order")
fam <- get_ra("family")
gen <- get_ra("genus")

taxa <- rbind(taxa, dom, king, phyl, class, ord, fam, gen)
```

get individual signatures for testing
```{r}
#689
frankelSig_up <- sub("^.+\\|", "", unlist(strsplit(sigs$MetaPhlAn.taxon.names[5],",")))

#679
matsonSig_up <- sub("^.+\\|", "", unlist(strsplit(c(sigs$MetaPhlAn.taxon.names[6],sigs$MetaPhlAn.taxon.names[7]), ",")))
matsonSig_down <- sub("^.+\\|", "", unlist(strsplit(sigs$MetaPhlAn.taxon.names[8],",")))

#688
gopaSig_up <- sub("^.+\\|", "", unlist(strsplit(sigs$MetaPhlAn.taxon.names[3] ,",")))
gopaSig_down <- sub("^.+\\|", "",unlist(strsplit(sigs$MetaPhlAn.taxon.names[4], ",")))

#691
windSig_up <- sub("^.+\\|", "", unlist(strsplit(sigs$MetaPhlAn.taxon.names[1] ,",")))
windSig_down <- sub("^.+\\|", "", unlist(strsplit(sigs$MetaPhlAn.taxon.names[2] ,",")))

paa_up <- c("s__Clostridium sp. UNK-MGS-6", "s__Firmicutes bacterium CAG 41", "s__Coprobacillus cateniformis", "s__Clostridia bacterium UC5-1-2G4","s__Streptococcus dysgalactiae", "s__Acidaminococcus sp. HPA0509", "s__Roseburia sp.CAG-303", "s__Butyrivibrio_sp. AE2032", "s__Coprobacillus sp. 29-1", "s__Coprobacillus sp. D6")

paa_down <- c("s__Alistipes sp. CAG-435", "s__Fusobacterium hwasookii", "s__Leuconostoc sp. DORA-2", "s__Paenibacillus lentimorbus", "s__Fusobacterium sp. oral taxon 370", "s__Alcanivorax hongdengensis", "s__Arthrobacter sp. LS16", "s__Bacillus sp. 5B6", "s__Bacillus velezensis", "s__Fusobacterium sp. OBRC1")
```


How do we handle a signature that only has up or down? Need to alter function to accommodate for both
could take an empty character vector, with checks for if empty

also need to consider handling if no microbes are present in the dataset

Calculate signature score
```{r}
# Function to calculate a microbe signature given a set of up and down regulated microbes and the relAbun of sample data
#
# @sig_up character vector listing microbes in given signature that are up-regulated
# @sig_down character vector listing microbes in given signature that are down-regulated
# @taxa df containing columns 'sample', 'microbe', 'exo.ra'
#
getMicroSigScore <- function(sig_up, sig_down, taxa){
  # z score transform relAbun 
  df <- taxa %>%
    group_by(sample) %>%
    mutate(z_score = scale(exo.ra))
  
  upMics <- df %>%
    filter(microbe %in% sig_up)
  
  upTotal <- length(unique(upMics$microbe))
  
  upMics <- upMics %>%  
    group_by(sample) %>%
    summarize(tempUp = sum(exo.ra))
  
  downMics <- df %>%
    filter(microbe %in% sig_down)
  
  downTotal <- length(unique(downMics$microbe))
  
  downMics <- downMics %>%
  group_by(sample) %>%
  summarize(tempDown = sum(exo.ra))
  
  if(upTotal == 0 && downTotal == 0){
    df <- data.frame(matrix(ncol = 0, nrow = 0))
  }else if(upTotal == 0){
    df <- downMics %>%
      mutate(score = tempDown*(1/downTotal)) %>%
      select(sample,score)
  }else if(downTotal == 0){
    df <- upMics %>%
      mutate(score = tempUp*(1/upTotal)) %>%
      select(sample, score)
  }else{
    df <- merge(upMics, downMics, by = "sample") %>%
    mutate(score = (tempUp - tempDown)*(1/(upTotal + downTotal))) %>%
    select(sample, score)
  }
  
  return(df)
}
```


get signature scores for sigs from literature
```{r}
wind <- getMicroSigScore(windSig_up, windSig_down, taxa) %>%
  rename(wind = score)

gopa <- getMicroSigScore(gopaSig_up, gopaSig_down, taxa) %>%
  rename(gopa = score)

# currently returns emtpy df - check presence of microbes
matson <- getMicroSigScore(matsonSig_up, matsonSig_down, taxa) %>%
  rename(matson = score)

frankel <- getMicroSigScore(frankelSig_up, c(), taxa) %>%
  rename(frankel = score)# pass and empty character vector for negatively correlated microbes 

paa <- getMicroSigScore(paa_up, paa_down, taxa) %>%
  rename(paa = score)
```

combine signatures into one table
```{r}
df_list <- list(wind, gopa, paa, frankel, matson)
microSigs <- df_list %>% reduce(full_join, by='sample')
```

<!-- testing area -->
<!-- ```{r} -->
<!-- unique (grep("taxon", taxa$microbe, value=TRUE)) -->
<!-- ``` -->

Test signature predictive power

######## Function to make the model and predictions ######
```{r}
getROC <- function(train, clin.train, test, clin.test){
  model.training <- randomForest(x = train[,-1, drop=FALSE], y = as.factor(clin.train$Response))
  test.preds <- predict(model.training, test[,-1, drop=FALSE])
  
  print(model.training)
  # Check variable importance
  varImpPlot(model.training)
  # Prediction confusion matrix
  table(observed = clin.test$Response, predicted = test.preds)
  
  prediction_for_roc_curve <- predict(model.training,test[,-1, drop=FALSE],type="prob")
  pred <- prediction(prediction_for_roc_curve[,2], clin.test$Response)
  perf <- performance(pred, "tpr", "fpr")
  
  auc_ROCR <- performance(pred, measure = "auc")
  print(auc_ROCR)
  auc <- auc_ROCR@y.values[[1]]
  print(auc)
  
  df <- data.frame(FalsePositive=c(perf@x.values[[1]]),
                    TruePositive=c(perf@y.values[[1]]))
  
  return(df)
}
```

Split into testing and training data
```{r}
testn <- round(.2 * nrow(response))
set.seed(12345)
testsamps <- sample(response$sample, testn)

# all clin vars
clin.train <- response %>%
  dplyr::filter(!sample %in% testsamps)
clin.test <- response %>%
  dplyr::filter(sample %in% testsamps)
```

### gopa testing and training ###
```{r}
train.exo <- microSigs %>%
  dplyr::filter(!sample %in% testsamps) %>%
  select(sample, gopa)
test.exo <- microSigs %>%
  dplyr::filter(sample %in% testsamps) %>%
  select(sample, gopa)
```

# model and plot ROC
```{r}
gopa_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

gopa_ra_plot <- ggplot(gopa_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="darkgreen")

gopa_ra_plot
```

wind testing and training
```{r}
train.exo <- microSigs %>%
  dplyr::filter(!sample %in% testsamps) %>%
  select(sample, wind)
test.exo <- microSigs %>%
  dplyr::filter(sample %in% testsamps) %>%
  select(sample, wind)
```

## model and plot ROC
```{r}
wind_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

wind_ra_plot <- ggplot(wind_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_smooth(se = FALSE, color="magenta")

wind_ra_plot
```

paa testing and training
```{r}
train.exo <- microSigs %>%
  dplyr::filter(!sample %in% testsamps) %>%
  select(sample, paa)
test.exo <- microSigs %>%
  dplyr::filter(sample %in% testsamps) %>%
  select(sample, paa)
```

## model and plot ROC
```{r}
paa_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

paa_ra_plot <- ggplot(paa_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  geom_line(color = "blue")
  #geom_smooth(se = FALSE, color="blue")

paa_ra_plot
```
### matson testing and training ###
```{r}
train.exo <- microSigs %>%
  dplyr::filter(!sample %in% testsamps) %>%
  select(sample, matson)
test.exo <- microSigs %>%
  dplyr::filter(sample %in% testsamps) %>%
  select(sample, matson)
```

## model and plot ROC
```{r}
mat_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

mat_ra_plot <- ggplot(mat_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  #geom_line(color = "blue")
  geom_smooth(se = FALSE, color="orange")

mat_ra_plot
```

### frankel testing and training ###
```{r}
train.exo <- microSigs %>%
  dplyr::filter(!sample %in% testsamps) %>%
  select(sample, frankel)
test.exo <- microSigs %>%
  dplyr::filter(sample %in% testsamps) %>%
  select(sample, frankel)
```

## model and plot ROC
```{r}
frank_ra <- getROC(train.exo, clin.train, test.exo, clin.test)

frank_ra_plot <- ggplot(frank_ra, aes(x=FalsePositive, y=TruePositive)) + 
  theme_bw() +
  #geom_line(color = "blue")
  geom_smooth(se = FALSE, color="purple")

frank_ra_plot
```

######################################
Combined ROC's
######################################
```{r}
temp <- combine(wind_ra, gopa_ra, paa_ra, mat_ra, frank_ra)

combined <- ggplot(temp, aes(x=FalsePositive, y=TruePositive, color=source)) + 
  theme_bw() +
  geom_line() +
  #geom_smooth(se = FALSE) +
  scale_color_manual(values=c("darkgoldenrod2", "cornflowerblue", "deeppink4", "darkgreen", "purple")) +
  xlab("Specifity") +
  ylab("Sensitivity") +
annotate(geom="text", x=0.75, y=0.3, label="AUC = 0.7",
            color="cornflowerblue", size=8) +
annotate(geom="text", x=0.75, y=0.2, label="AUC = 0.8",
            color="darkgoldenrod2", size=8) +
  annotate(geom="text", x=0.75, y=0.1, label="AUC = 0.95",
            color="deeppink4", size=8) +
  annotate(geom="text", x=0.75, y=0.5, label="AUC = 0.55",
            color="darkgreen", size=8) +
  annotate(geom="text", x=0.75, y=0.4, label="AUC = 0.6875",
            color="purple", size=8) 

combined

ggsave(filename = "../figures/microSig_combined_ROC.png", plot = combined, width = 28, height = 10, dpi = 400)
```

