---
title: "immune cell"
author: "Caroline Wheeler"
date: "8/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(reshape2)
```

read in data
```{r}
clin <- read.table("/fs/ess/PAS1695/projects/exorien/data/IO_NOVA_clin.txt", header = TRUE) %>%
  rename("sample" = Sample_ID) %>%
  filter(Cancer_Type == "Melanoma") %>%
  select(-Cancer_Type, -OS_Event, -OS)

cells <- read.csv("/fs/ess/PAS1695/projects/exorien/data/cibersort/2022-03-16_immunecell_composition.csv") %>%
   dplyr::select(-X, -P.value, -Correlation, -RMSE)
```

```{r}
cells <- merge(cells, clin)
```

kruskal-wallis test
```{r}
results <- list()
for(i in colnames(cells[,2:23])){  
  results[[i]] <- kruskal.test(formula(paste(i, "~ Response")), data = cells)
}

results 
```

T.cells.CD8 by Response
Kruskal-Wallis chi-squared = 9.131, df = 1, p-value = 0.002513

T.cells.CD4.memory.activated by Response
Kruskal-Wallis chi-squared = 9.6648, df = 1, p-value = 0.001878

 NK.cells.activated by Response
Kruskal-Wallis chi-squared = 4.5242, df = 1, p-value = 0.03342

Macrophages.M1 by Response
Kruskal-Wallis chi-squared = 6.4522, df = 1, p-value = 0.01108

Mast.cells.resting by Response
Kruskal-Wallis chi-squared = 6.1322, df = 1, p-value = 0.01327


```{r}
cells <- cells %>%
  select(-sample)

collabs <- c("Naive B-cells", "Memory B-cells", "Plasma cells", "CD8+ T-cells", "Naive CD4 T-cells", "Memory CD4 T-cells, resting", "Memory CD4 T-cells, activated", "Follicular-helper T-cells", "Tregs", "Gamma delta T-cells", "NK cells, resting", "NK cells, activated", "Monocytes", "M0 Macrophages", "M1 Macrophages", "M2 Macrophages", "Dendritic cells, resting", "Dendritic cells, activated", "Mast cells, resting", "Mast cells, activated", "Eosinophils", "Neutrophils", "Response")

colnames(cells) <- collabs

cells <- melt(cells, id.vars= 23)
cells$Response[cells$Response == 1] <- "Responder"
cells$Response[cells$Response == 0] <- "Non-Responder"
```


```{r}
bp <- ggplot(cells, aes(x=variable, y=value)) + geom_boxplot(aes(fill=Response)) + xlab("") + ylab("Abundance") + labs(fill = "") + theme_bw(base_size = 28) + theme(axis.text.x = element_text(angle = 45, hjust=1)) + scale_fill_manual(values=c("red3", "cornflowerblue"))
  
bp
```

save plot
```{r}
ggsave(filename = "../figures/immunecell_boxplot.png", plot = bp, width = 28, height = 10, dpi = 600)
```
