---
title: "bugsigdb"
author: "Caroline Wheeler"
date: "12/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(bugsigdbr)
library(tidyverse)
library(ontologyIndex)
```

```{r}
bsdb <- importBugSigDB()
colnames(bsdb)
conds <- as.data.frame(table(bsdb$Condition))
sites <- as.data.frame(table(bsdb$`Body site`))
```
Subsetting
```{r}
obese <- bsdb %>%
  filter(Condition == "obesity",
         `Host species` == "Homo sapiens")

colrec <- bsdb %>%
  filter(Condition == "colorectal cancer",
          `Host species` == "Homo sapiens")
```

Get signatures
```{r}
sigs <- getSignatures(colrec)

# signatures based on the full taxonomic classification in MetaPhlAn format
mp.sigs <- getSignatures(colrec, tax.id.type = "metaphlan")

#taxonmic name only
tn.sigs <- getSignatures(colrec, tax.id.type = "taxname")

#genus level
gn.sigs <- getSignatures(colrec, 
                         tax.id.type = "taxname",
                         tax.level = "genus",
                         exact.tax.level = FALSE)
```

Ontology    
```{r}
efo <- getOntology("efo")
cancer <- subsetByOntology(bsdb,
                        column = "Condition",
                        term = "cancer",
                        ontology = efo)

table(cancer[,"Condition"])

mp.sigs.cancer <- getSignatures(cancer, tax.id.type = "metaphlan")
```

```{r}
colrec_sigs <- plyr::ldply(gn.sigs, rbind)
```


### IO response group ####
```{r}
io <- cancer %>%
  filter(`Group 1 name` == "Responders" | `Group 1 name` == "Progressive -Free- Survival more than 30 days")

io_mics <- getSignatures(io, tax.id.type = "taxname")
io_mics <- plyr::ldply(io_mics, rbind)

io_mics_met <- getSignatures(io, tax.id.type = "metaphlan")
io_mics_met <- plyr::ldply(io_mics_met, rbind)
```

Why is Prevotella in both??

Responders:
Akkermansia muciniphila
Eubacterium sp.
Lachnospiraceae
Erysipelotrichaceae
Intestinimonas
Alistipes finegoldii
Bacteroides sp.
Cloacibacillus porcorum
Enterococcus faecium
Prevotella
Eubacterium sp.
Alistipes
Akkermansia muciniphila
Intestinimonas
Bacteroides nordii
Bacteroides xylanisolvens
Blautia
Lachnospiraceae
Firmicutes
Eubacteriales
Ruminococcus
Clostridium sp.
Flavonifractor plautii
Bacteroides caccae
Ruminococcus sp.

Non-Responders:
Prevotella
Clostridium sp.
Eubacteriales
Enterocloster bolteae
Blautia
Bacteroides clarus
Proteobacteria
Bacteroides nordii
Parabacteroides distasonis
Staphylococcus haemolyticus
Corynebacterium aurimucosum
Parabacteroides distasonis
Eubacteriales
Anaerotruncus colihominis
Lachnospiraceae
Erysipelotrichaceae

Met melanoma and more sigs available on bugsigdbr website....




